---
title: "Data Visualization with ggplot2 (Part 1)"
author: "MOHAMMAD SHADAN"
date: "July 15, 2017"
output: 
  html_document: 
    keep_md: yes
---

# 1. Introduction

## Introduction - Video

### Explore and Explain

In this video we made the distinction between plots for exploring and plots for explaining data. Which of the following are exploratory plots typically NOT?

Possible Answers

 - Meant for a specialist audience.
 - Data-heavy.
 - Pretty. (Incorrect)
 - Rough first drafts.
 - Part of our data science toolkit as graphical data analysis.
 
. You're not concerned with beautiful at this point. Although, the plots should be meaningful and conform to best practices so that you are not misled!

### Exploring ggplot2, part 1

To get a first feel for ggplot2, let's try to run some basic ggplot2 commands. Together, they build a plot of the mtcars dataset that contains information about 32 cars from a 1973 Motor Trend magazine. This dataset is small, intuitive, and contains a variety of continuous and categorical variables.

Instructions
 - Load the ggplot2 package using library(). It is already installed on DataCamp's servers.
 - Use str() to explore the structure of the mtcars dataset.
 - Hit Submit Answer. This will execute the example code on the right. See if you can understand what ggplot does with the data.
 
```{r}
# Load the ggplot2 package
library(ggplot2)

# Explore the mtcars data frame with str()
str(mtcars)

# Execute the following command
ggplot(mtcars, aes(x = cyl, y = mpg)) +
  geom_point()
```

Notice that ggplot2 treats cyl as a continuous variable. You get a plot, but it's not quite right, because it gives the impression that there is such a thing as a 5 or 7-cylinder car, which there is not.

### Exploring ggplot2, part 2

The plot from the previous exercise wasn't really satisfying. Although cyl (the number of cylinders) is categorical, it is classified as numeric in mtcars. You'll have to explicitly tell ggplot2 that cyl is a categorical variable.

Instructions
 - Change the ggplot() command by wrapping factor() around cyl.
 - Hit Submit Answer and see if the resulting plot is better this time.

```{r}
# Load the ggplot2 package
library(ggplot2)

# Change the command below so that cyl is treated as factor
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_point()
```

Notice that ggplot2 treats cyl as a factor. This time the x-axis does not contain variables like 5 or 7, only the values that are present in the dataset.

## Grammar of Graphics - Video

### Exploring ggplot2, part 3

We'll use several datasets throughout the courses to showcase the concepts discussed in the videos. In the previous exercises, you already got to know mtcars. Let's dive a little deeper to explore the three main topics in this course: The data, aesthetics, and geom layers.

The mtcars dataset contains information about 32 cars from 1973 Motor Trend magazine. This dataset is small, intuitive, and contains a variety of continuous and categorical variables.

You're encouraged to think about how the examples and concepts we discuss throughout these data viz courses apply to your own data-sets!

Instructions

 - ggplot2 has already been loaded for you. Take a look at the first command. It plots the mpg (miles per galon) against the weight (in thousands of pounds). You don't have to change anything about this command.
 - In the second call of ggplot() change the color argument in aes() (which stands for aesthetics). The color should be dependent on the displacement of the car engine, found in disp.
 - In the third call of ggplot() change the size argument in aes() (which stands for aesthetics). The size should be dependent on the displacement of the car engine, found in disp.
 
```{r}
# A scatter plot has been made for you
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# Replace ___ with the correct vector
ggplot(mtcars, aes(x = wt, y = mpg, col = disp)) +
  geom_point()

# Replace ___ with the correct vector
ggplot(mtcars, aes(x = wt, y = mpg, size = disp)) +
  geom_point()

```

Notice that a legend for the color and size scales was automatically generated.

### Understanding Variables

In the previous exercise you saw that disp can be mapped onto a color gradient or onto a continuous size scale.

Another argument of aes() is the shape of the points. There are a finite number of shapes which ggplot() can automatically assign to the points. However, if you try this command in the console to the right:
  
```{r}
ggplot(mtcars, aes(x = wt, y = mpg, shape = disp)) +
  geom_point()
```
It gives an error. What does this mean?

Possible Answers

 - shape is not a defined argument.
 - shape only makes sense with categorical data, and disp is continuous. (Correct)
 - shape only makes sense with continuous data, and disp is categorical.
 - shape is not a variable in your dataset.
 - shape has to be defined as a function.

The error message 'A continuous variable can not be mapped to shape', means that shape doesn't exist on a continuous scale here.

## ggplot2 - Video

### Exploring ggplot2, part 4

The diamonds data frame contains information on the prices and various metrics of 50,000 diamonds. Among the variables included are carat (a measurement of the size of the diamond) and price. For the next exercises, you'll be using a subset of 1,000 diamonds.

Here you'll use two common geom layer functions: geom_point() and geom_smooth(). We already saw in the earlier exercises how these are added using the + operator.

Instructions
 - Explore the diamonds data frame with the str() function.
 - Use the + operator to add geom_point() to the first ggplot() command. This will tell ggplot2 to draw points on the plot.
 - Use the + operator to add geom_point() and geom_smooth(). These just stack on each other! geom_smooth() will draw a smoothed line over the points.
 
```{r}
# Explore the diamonds data frame with str()
str(diamonds)

# Add geom_point() with +
ggplot(diamonds, aes(x = carat, y = price)) + geom_point()


# Add geom_point() and geom_smooth() with +
ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + geom_smooth()
```
If you had executed the command without adding a +, it would produce an error message 'No layers in plot' because you are missing the third essential layer - the geom layer.

### Exploring ggplot2, part 5

The code for last plot of the previous exercise is available in the script on the right. It builds a scatter plot of the diamonds dataset, with carat on the x-axis and price on the y-axis. geom_smooth() is used to add a smooth line.

With this plot as a starting point, let's explore some more possibilities of combining geoms.

Instructions
 - Plot 2 - Copy and paste plot 1, but show only the smooth line, no points.
 - Plot 3 - Show only the smooth line, but color according to clarity by placing the argument color = clarity in the aes() function of your ggplot() call.
 - Plot 4 - Draw translucent colored points.
Copy the ggplot() command from plot 3 (with clarity mapped to color).
 - Remove the smooth layer.
 - Add the points layer back in.
 - Set alpha = 0.4 inside geom_point(). This will make the points 40% transparent.
 
```{r}
# 1 - The plot you created in the previous exercise
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() +
  geom_smooth()

# 2 - Copy the above command but show only the smooth line
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_smooth()


# 3 - Copy the above command and assign the correct value to col in aes()
ggplot(diamonds, aes(x = carat, y = price, color=clarity)) +
  geom_smooth()


# 4 - Keep the color settings from previous command. Plot only the points with argument alpha.
ggplot(diamonds, aes(x = carat, y = price, col=clarity)) +
  geom_point(alpha=0.4)

```
geom_point() + geom_smooth() is a common combination.

### Understanding the grammar, part 1

Here you'll explore some of the different grammatical elements. Throughout this course, you'll discover how they can be combined in all sorts of ways to develop unique plots.

In the following instructions, you'll start by creating a ggplot object from the diamonds dataset. Next, you'll add layers onto this object to build beautiful & informative plots.

Instructions
 - Define the data (diamonds) and aesthetics layers. Map carat on the x axis and price on the y axis. Assign it to an object: dia_plot.
 - Using +, add a geom_point() layer (with no arguments), to the dia_plot object.   - This can be in a single or multiple lines.
 - Note that you can also call aes() within the geom_point() function. Map clarity to the color argument in this way.


```{r}
# Create the object containing the data and aes layers: dia_plot
dia_plot <- ggplot(diamonds, aes(x = carat, y = price))

# Add a geom layer with + and geom_point()
dia_plot + geom_point()

# Add the same geom layer, but with aes() inside
dia_plot + geom_point(aes(color = clarity))

```

Notice how you can store the plot as a ggplot object that you can use later on to add other layers; that's pretty convenient!

### Understanding the grammar, part 2

Continuing with the previous exercise, here you'll explore mixing arguments and aesthetics in a single geometry.

You're still working on the diamonds dataset.

Instructions
 - 1 - The dia_plot object has been created for you.
 - 2 - Update dia_plot so that it contains all the functions to make a scatter plot by using geom_point() for the geom layer. Set alpha = 0.2.
 - 3 - Using +, plot the dia_plot object with a geom_smooth() layer on top. You don't want any error shading, which can be achieved by setting the se = FALSE in geom_smooth().
 - 4 - Modify the geom_smooth() function from the previous instruction so that it contains aes() and map clarity to the col argument. 

```{r}
# 1 - The dia_plot object has been created for you
dia_plot <- ggplot(diamonds, aes(x = carat, y = price))

# 2 - Expand dia_plot by adding geom_point() with alpha set to 0.2
dia_plot <- dia_plot + geom_point(alpha=0.2)

# 3 - Plot dia_plot with additional geom_smooth() with se set to FALSE
dia_plot + geom_smooth(se = FALSE)

# 4 - Copy the command from above and add aes() with the correct mapping to geom_smooth()
dia_plot + geom_smooth(aes(col = clarity), se = FALSE)
```

To set a property of a geom to a single value, pass it as an argument. To give the property different values for each row of data, pass it as an aesthetic.